SMT-Based Context-Bounded Model Checking for Embedded Systems: Challenges and Future Trends

### 2016-03-26 paper.tex
corrections to section 2

Bounded Model Checking (BMC) based on Boolean Satisfiability (SAT) was originally proposed to verify hardware designs~\cite{Biere99,handbook09}. Indeed, Biere {\it et al.} were able to successfully verify large digital circuits with approximately $9510$ latches and $9499$ inputs, leading to BMC formulae with $4 \times 10^6$ variables and $1.2 \times 10^7$ clauses to be checked by a standard SAT solver. BMC based on Satisfiability Modulo Theories (SMT)~\cite{BarrettSST09} was originally proposed to deal with increasing software verification complexity~\cite{Armando06}.

In general, BMC techniques aim to check the violation of a given (safety) property at a given system's depth, {\it i.e.}, given a transition system \textit{M}, which is derived from the control-flow graph of a program; a property $\phi$ that represents the program correctness and/or the system's behaviour; and a bound of iterations \textit{k}, which limits the loop unrolling, data structures, and context-switches, BMC techniques thus unfold the system \textit{k} times, in order to convert it into a verification condition $\psi$, expressed in propositional logic or in a decidable-fragment of first-order logic, such that $\psi$ is \textit{satisfiable} if and only if $\phi$ has a counterexample of depth less than or equal to \textit{k}.

From the practical point of view, SAT-based or SMT-based BMC procedures have been successfully applied to verify a large number of hardware and software systems, including digital circuits, single- and multi-threaded programs. Those BMC techniques were able to find subtle bugs in real digital and embedded software systems, as reported in the literature~\cite{Clarke04,MerzFS12,CordeiroF11,Ivancic05,Cordeiro12}. However, the main criticism with respect to BMC techniques relies on completeness since they are able to prove system correctness only if an upper bound \textit{k} is known, {\it i.e.}, a bound that unfolds all loops and recursive functions to their maximum possible depth.

Due to that limitation, BMC tools are typically susceptible to exhaustion of time or memory limits, when checking complex circuits implementation or programs with loops whose bounds are too large or cannot be statically determined.

@@

One promising approach to achieve completeness in BMC techniques is to prove that an invariant (assertion) is \textit{k}-inductive~\cite{EenS03,Sheera00}. The main challenge, however, of such approach relies on computing and strengthening inductive invariants from the program. In particular, loop invariants computed from the program must be inductive (and not just invariant) for the corresponding verification conditions to be valid, {\it i.e.}, induction cannot determine the invariance of a non-inductive assertion~\cite{Bradley07}. As a consequence, even if \textit{k}-induction procedures successfully compute such assertions, which are indeed invariant, those must be inductive so that verifiers can automatically prove program correctness.

There are several invariant-generation algorithms that discover linear and polynomial relations among integer and real variables, in order to provide loop invariants and also to discover the memory ``shape'' in programming languages with pointers~\cite{pips:2013,Henry:2012}. The current literature also reports successful applications of \textit{k}-induction based verification algorithms for hardware and software systems using invariant generation and strengthening, mostly based on interval analysis.

Novel verification algorithms for proving correctness of (a large set of) C programs, by mathematical induction, in a completely automatic way ({\it i.e.}, users do not need to provide the loop invariant) were proposed~\cite{Gadelha15,Beyer15,Brain15,Rocha15}. Additionally, \textit{k}-induction based verification was also applied to ensure that (restricted) C programs (1) do not contain violations related to data races~\cite{Donaldson10,Kinductor} considering the  Cell BE processor and (2) do respect time constraints, which are specified during the system design phase~\cite{EenS03}. Apart from that, the \textit{k}-induction algorithm is also a well-established technique in hardware verification, where it is easily applied, due to the monolithic transition relation present in such designs~\cite{EenS03,Sheera00,GrosseLD09}.

@@

It is worth noticing that, currently, SMT-based BMC approaches check code properties in real programs, which basically address programing-language issues and general correctness, without taking into account target applications or system behavior. Such a statement is important, since, as already mentioned, many system features are being moved to software domain, which then requires schemes that they not only check if the source code is correctly written, but also if it will properly respond in real environments or under external problems. For instance, the anti-lock braking system software of a vehicle model can be bug free, but it may not work correctly if a sensor is damaged.

Indeed, research in software verification is now taking such considerations into account and some schemes already use knowledge about the system to be verified and the underlying hardware. Recently, a verification tool proposed for digital systems, called digital system verifier (DSVerifier)~\cite{dsv_spin2015}, which is able to aid engineers to check overflow, limit cycle, output error, timing, stability, and minimum phase, considering finite word-length effects. Indeed, DSVerifier is a useful test tool, which takes into account realization forms ({\it e.g.}, direct forms, delta forms, and transposed forms) and other implementation restrictions to explore the design space. Ultimately, if the system requirements are not met with a given configuration, an analysis of the provided error report may suggest another setup.

Scratch is another example of a software model checker, which uses knowledge about the system to be verified and the underlying hardware, for detecting races related to direct memory access (DMA) in the Cell BE processor~\cite{Donaldson10}. The tool also uses BMC, in order to detect DMA races, and BMC with \textit{k}-induction, which aims to prove the absence of races.

>>>>

Bounded Model Checking (BMC) based on Boolean Satisfiability (SAT) was originally proposed to verify hardware designs~\cite{Biere99,handbook09}. Indeed, Biere {\it et al.} were able to successfully verify large digital circuits with approximately $9510$ latches and $9499$ inputs, leading to BMC formulae with $4 \times 10^6$ variables and $1.2 \times 10^7$ clauses to be checked by a standard SAT solver. BMC based on Satisfiability Modulo Theories (SMT)~\cite{BarrettSST09}, in turn, was originally proposed to deal with increasing software verification complexity~\cite{Armando06}.

In general, BMC techniques aim to check the violation of a given (safety) property at a given system depth. Indeed, given a transition system \textit{M}, which is derived from the control-flow graph of a program, a property $\phi$, which represents the program correctness and/or the system's behavior, and an iteration bound \textit{k}, which limits the loop unrolling, data structures, and context-switches, BMC techniques thus unfold the system \textit{k} times, in order to convert it into a verification condition $\psi$, expressed in propositional logic or in a decidable-fragment of first-order logic, such that $\psi$ is \textit{satisfiable} if and only if $\phi$ has a counterexample of depth less than or equal to \textit{k}.

From the practical point of view, SAT-based or SMT-based BMC procedures have been successfully applied to verify a large number of hardware and software systems, including digital circuits and single- and multi-threaded programs. Those BMC techniques were able to find subtle bugs in real digital and embedded software systems, as reported in the available literature~\cite{Clarke04,MerzFS12,CordeiroF11,Ivancic05,Cordeiro12}. Nonetheless, the main criticism with respect to BMC techniques relies on completeness, since they are able to prove system correctness only if an upper bound \textit{k} is known, {\it i.e.}, a bound that unfolds all loops and recursive functions to their maximum possible depth.

Due to that limitation, BMC tools are typically susceptible to exhaustion of time or memory limits, when checking complex circuit-implementations or programs with loops, whose bounds are too large or cannot be statically determined.

@@

One promising approach to achieve completeness, in BMC techniques, is to prove that an invariant (assertion) is \textit{k}-inductive~\cite{EenS03,Sheera00}. The main challenge, however, of such an approach relies on computing and strengthening inductive invariants from programs. In particular, loop invariants, which were computed from programs under verifications, must be inductive (and not just invariant) for the corresponding verification conditions to be valid, {\it i.e.}, induction cannot determine the invariance of a non-inductive assertion~\cite{Bradley07}. As a consequence, even if \textit{k}-induction procedures successfully compute such assertions, which are indeed invariant, those must be inductive, so that verifiers can automatically prove program correctness.

There are several invariant-generation algorithms that discover linear and polynomial relations among integer and real variables, in order to provide loop invariants and also discover the memory ``shape'', in programming languages with pointers~\cite{pips:2013,Henry:2012}. The current literature also reports successful applications of \textit{k}-induction based verification algorithms for hardware and software systems, using invariant generation and strengthening, mostly based on interval analysis.

Novel verification algorithms for proving correctness of (a large set of) C programs, by mathematical induction, in a completely automatic way ({\it i.e.}, users do not need to provide the loop invariant) were recently proposed~\cite{Gadelha15,Beyer15,Brain15,Rocha15}. Additionally, \textit{k}-induction based verification was also applied to ensure that (restricted) C programs (1) do not contain violations related to data races~\cite{Donaldson10,Kinductor}, considering the Cell BE processor, and (2) do respect time constraints, which are specified during the system design phase~\cite{EenS03}. Apart from that, the \textit{k}-induction algorithm is also a well-established technique in hardware verification, where it is easily applied, due to the monolithic transition relation present in such designs~\cite{EenS03,Sheera00,GrosseLD09}.

@@

It is worth noticing that, currently, SMT-based BMC approaches check code properties in real programs, which basically address programing-language issues and general correctness, without taking into account target applications or system behavior. Such a statement is important, since, as already mentioned, many system features are being moved to software domain, which then requires schemes that do not only check if the source code is correctly written, but also if it will properly respond in real environments or under external problems. For instance, the anti-lock braking system software of a vehicle model can be bug free, but it may not work correctly if a sensor is damaged.

Indeed, research in software verification is now incorporating such considerations, during checking processes, and some schemes already use knowledge about the system to be verified and the underlying hardware. Recently, a verification tool for digital systems was proposed, which is called digital system verifier (DSVerifier)~\cite{dsv_spin2015} and is able to aid engineers to check overflow, limit cycle, output error, timing, stability, and minimum phase, considering finite word-length effects. Indeed, DSVerifier is a useful test tool, which takes into account realization forms ({\it e.g.}, direct forms, delta forms, and transposed forms) and other implementation restrictions to explore the design space. Ultimately, if the system requirements are not met with a given configuration, an analysis of the provided error report may suggest another setup.

Scratch is another example of a software model checker, which uses knowledge about the system to be verified and the underlying hardware, for detecting races related to direct memory access (DMA), in the Cell BE processor~\cite{Donaldson10}. That tool also uses BMC, in order to detect DMA races, and BMC with \textit{k}-induction, which aims to prove the absence of races. If support to other DMA operations were added, Scratch could be adapted to different architectures, {\it i.e.}, the same techniques would be employed, but with a different system behavior/knowledge.
